---
title: "DESeq2 Analysis Pipeline"
author: "Bulk RNA-seq"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
---

# DESeq2 Differential Expression Analysis Pipeline

A universal, modular pipeline for DESeq2 analysis with volcano plots, heatmaps, and PCA.

## 1. Setup & Configuration

```{r setup, message=FALSE, warning=FALSE}


# Activate renv for this project (creates renv structure if needed)
# Renv will manage package versions and dependencies for reproducibility
library(renv)
renv::activate()

# Restore packages from lockfile (pulls from cache if available)
if (file.exists("renv.lock")) {
  renv::restore(prompt = FALSE)
}
packages <- c(
  # CRAN
  "dplyr", "stringr", "readr", "ggplot2", "ggrepel",
  "pheatmap", "systemfonts", "AsioHeaders", "httpgd",
  # Bioconductor
  "rhdf5", "DESeq2", "tximport", "AnnotationDbi",
  "org.Mm.eg.db", "GenomicFeatures", "EnhancedVolcano"
)

#Install any missing packages
installed <- packages %in% rownames(installed.packages())
if (any(!installed)) {
  renv::install(packages[!installed], prompt = FALSE)
}

#Install httpgd for remote figure generation for VSCode if using
if (!"httpgd" %in% rownames(installed.packages())) {
  renv::install("httpgd", repos = "https://community.r-multiverse.org")
}

# Load all packages
invisible(lapply(c(packages, "httpgd"), function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))

#Save the state of the environment (creates renv.lock)
renv::snapshot(prompt = FALSE)

knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Configuration: User Inputs

```{r config}
# ===== EDIT THESE PARAMETERS =====
# Paths
SAMPLE_PATH <- "/projects/margeta/phagocytosis/kallisto_results/mouse_ensbl/samples.txt"
GTF_PATH <- "/data_pool/raw_data/reference_genomes/mouse/ensembl/Mus_musculus.GRCm39.115.gtf"
TX2GENE_PATH <- "/data_pool/raw_data/reference_genomes/mouse/ensembl/tx2gene.tsv"
OUTPUT_DIR <- "/projects/margeta/phagocytosis/kallisto_results/mouse_ensbl/results"

# Annotation
ORGANISM <- org.Mm.eg.db
ORGANISM_NAME <- "Mus musculus"

# Gene filtering (remove NA gene symbols and patterns like "Gm...")
FILTER_NA_SYMBOL <- TRUE
FILTER_PATTERN <- "^Gm"  # Set to "" to skip pattern filtering

# Volcano plot configuration
LABEL_GENES <- c("Csta3", "Trem2", "Gpnmb", "Cstb", "Spp1", "Lgals3", "Apoe", "Lgals1")
VOLCANO_PADJ <- 0.05
VOLCANO_LFC <- 1

# Heatmap configuration - genes to highlight
HEATMAP_PATH_GENES <- c(
  "Csta3", "Trem2", "Gpnmb", "Cstb", "Spp1", "Lgals3", "Apoe", "Lgals1",
  "C1qb", "C1qc", "Ctsd", "Ctsa", "Hexa", "Ifi27", "Ifi35", "Siglecg",
  "CD74", "H2-Aa"
)

# Top genes heatmap (show top N up- and down-regulated)
TOP_GENES_N <- 25

# ===== END OF USER EDITS =====
dir.create(OUTPUT_DIR, showWarnings = FALSE, recursive = TRUE)
```
## 2. Load Tx2Gene (Transcript to Gene) Mapping

```{r tx2gene}
# Create or load tx2gene mapping (ENSMBL transcript ID → gene ID)
create_tx2gene <- function(gtf_path, output_path) {
  message("Creating tx2gene mapping from GTF...")
  txdb <- makeTxDbFromGFF(gtf_path)
  k <- keys(txdb, keytype = "TXNAME")
  tx2gene <- select(txdb, keys = k,
                    columns = "GENEID",
                    keytype = "TXNAME")
  # Remove version numbers from transcript IDs
  tx2gene[, 1] <- sub("\\..*$", "", tx2gene[, 1])
  write.table(tx2gene,
              file = output_path,
              quote = FALSE,
              sep = "\t",
              row.names = FALSE,
              col.names = FALSE)
  return(tx2gene)
}

if (!file.exists(TX2GENE_PATH)) {
  message(sprintf("Creating tx2gene at: %s", TX2GENE_PATH))
  tx2gene <- create_tx2gene(GTF_PATH, TX2GENE_PATH)
} else {
  message(sprintf("Loading tx2gene from: %s", TX2GENE_PATH))
  tx2gene <- read_tsv(TX2GENE_PATH, col_names = c("TXNAME", "GENEID"))
}

head(tx2gene)
```
## 3. Import Kallisto Results & Create DESeq2 Object

```{r deseq_import}
# Import kallisto counts via tximport
import_kallisto <- function(sample_path, tx2gene) {
  message("Reading sample metadata from:", sample_path)
  samples <- read.table(sample_path, header = TRUE, sep = ",")

  if (!all(c("sample", "path", "condition") %in% colnames(samples))) {
    stop("Sample file must have columns: sample, path, condition")
  }

  message("Importing kallisto results from", nrow(samples), "samples...")
  # Build named vector of kallisto abundance files
  files <- setNames(samples$path, samples$sample)

  # Check that files exist
  missing <- !file.exists(files)
  if (any(missing)) {
    stop("Missing files:\n", paste(names(files)[missing], collapse = "\n"))
  }

  # Import kallisto results
  txi <- tximport(
    files,
    type = "kallisto",
    tx2gene = tx2gene,
    countsFromAbundance = "no",
    ignoreTxVersion = TRUE
  )

  # Build colData (metadata)
  coldata <- as.data.frame(samples[, c("condition", "replicate", "sample"),
                                   drop = FALSE])
  rownames(coldata) <- samples$sample

  # Remove duplicate columns
  coldata <- coldata[, !duplicated(colnames(coldata))]

  # Build DESeqDataSet
  message("Creating DESeqDataSet...")
  dds <- DESeq2::DESeqDataSetFromTximport(
    txi,
    colData = coldata,
    design = ~ condition
  )

  dds
}

# Import data
dds <- import_kallisto(SAMPLE_PATH, tx2gene)
message("DESeqDataSet created: ", nrow(dds), " genes x ", ncol(dds), " samples")
```
## 4. Run Differential Expression Analysis

```{r deseq_analysis}
# Run DESeq2
message("Running DESeq2...")
dds <- DESeq(dds)

# Extract and annotate results
res <- results(dds)
res <- res[order(res$padj), ]

# Add gene symbols
res$symbol <- mapIds(
  ORGANISM,
  keys = rownames(res),
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
)

# Add to rowData for later use
rowData(dds)$symbol <- res$symbol

message("DESeq2 complete: ", sum(res$padj < 0.05, na.rm = TRUE), 
        " significant genes (padj < 0.05)")

# Summary statistics
summary(res)

# Save results to CSV
results_file <- file.path(OUTPUT_DIR, "deseq2_results_all.csv")
write.csv(as.data.frame(res), file = results_file)
message("Results saved to:", results_file)
```


## 5. Volcano Plot

```{r volcano, fig.width = 11, fig.height = 8}
# Function to create EnhancedVolcano plot
make_volcano <- function(res, label_genes = NULL, 
                         padj_cutoff = 0.05, lfc_cutoff = 1,
                         output_path = NULL) {
  res_df <- as.data.frame(res)
  res_df <- res_df %>% filter(!is.na(symbol))

  p <- EnhancedVolcano(
    res_df,
    lab = res_df$symbol,
    x = "log2FoldChange",
    y = "padj",
    ylim = c(0, -log10(10e-5)),
    xlim = c(-12, 12),
    selectLab = label_genes,
    labCol = "black",            # label color
    pCutoff = padj_cutoff,
    FCcutoff = lfc_cutoff,
    pointSize = 0.5,
    labSize = 6.0,
    colConnectors = "grey30",
    gridlines.major = FALSE,
    gridlines.minor = FALSE,
    col = c("gray", "gray", "gray", "red3"),
    colAlpha = 0.8,
    drawConnectors = TRUE,
    widthConnectors = 0.5,
    titleLabSize = 14,
    axisLabSize = 12
  )

  if (!is.null(output_path)) {
    message("Saving volcano plot to:", output_path)
    ggplot2::ggsave(output_path, p, width = 10, height = 15, dpi = 300)
  }
  p
}

# Create volcano plot
p_volcano <- make_volcano(
  res,
  label_genes = LABEL_GENES,
  padj_cutoff = VOLCANO_PADJ,
  lfc_cutoff = VOLCANO_LFC,
  output_path = file.path(OUTPUT_DIR, "volcano_plot.png")
)
p_volcano
```

## 6. Heatmap: Selected Gene Pathway

```{r heatmap_pathway, fig.width = 10, fig.height = 8}
# Heatmap of user-selected pathway genes
make_pathway_heatmap <- function(dds, genes_of_interest, output_path = NULL) {
  # Normalize counts
  norm_counts <- counts(dds, normalized = TRUE)
  rownames(norm_counts) <- AnnotationDbi::mapIds(
    ORGANISM,
    keys = rownames(norm_counts),
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )

  # Subset for selected genes (keep only those that exist)
  genes_present <- genes_of_interest[genes_of_interest %in% rownames(norm_counts)]
  missing <- genes_of_interest[!genes_of_interest %in% rownames(norm_counts)]

  if (length(missing) > 0) {
    message("Warning: genes not found:", paste(missing, collapse = ", "))
  }

  mat <- norm_counts[genes_present, ]
  # Log-transform and scale
  mat_log <- log2(mat + 1)
  mat_scaled <- t(scale(t(mat_log)))

  # Color palette
  colors <- colorRampPalette(c("navy", "white", "firebrick3"))(100)

  # Heatmap
  p <- pheatmap(
    mat_scaled,
    color = colors,
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    show_rownames = TRUE,
    show_colnames = TRUE,
    fontsize = 11,
    border_color = NA,
    main = "Selected Genes (Z-score normalized)",
    cellwidth = 25,
    cellheight = 15
  )

  if (!is.null(output_path)) {
    message("Saving pathway heatmap to:", output_path)
    pdf(output_path, width = 10, height = 8)
    print(p)
    dev.off()
  }

  p
}

# Create pathway heatmap
p_pathway <- make_pathway_heatmap(
  dds,
  genes_of_interest = HEATMAP_PATH_GENES,
  output_path = file.path(OUTPUT_DIR, "heatmap_pathway.pdf")
)
```
## 7. Heatmap: Top Differentially Expressed Genes

```{r heatmap_top, fig.width = 10, fig.height = 12}
# Heatmap of top N up- and down-regulated genes
make_top_genes_heatmap <- function(res, dds, n = 25,
                                   filter_genes = HEATMAP_PATH_GENES,
                                   output_path = NULL) {
  # Filter results
  res_filt <- as.data.frame(res)
  res_filt <- res_filt[!is.na(res_filt$log2FoldChange), ]

  if (FILTER_NA_SYMBOL) {
    res_filt <- res_filt[!is.na(res_filt$symbol), ]
  }

  if (FILTER_PATTERN != "") {
    res_filt <- res_filt[!grepl(FILTER_PATTERN, res_filt$symbol,
                                ignore.case = TRUE), ]
  }

  # Select top n up and n down
  res_ordered <- res_filt[order(res_filt$log2FoldChange, decreasing = TRUE), ]
  top_up <- head(res_ordered, n)
  top_down <- tail(res_ordered, n)
  top_genes <- unique(c(rownames(top_up), rownames(top_down)))

  # Get normalized counts and log-transform
  dds_sub <- assay(dds[top_genes, ])
  rownames(dds_sub) <- mapIds(
    ORGANISM,
    keys = rownames(dds_sub),
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )

  dds_sub <- log2(dds_sub + 1)

  # scale by row so each gene shows relative expression
  dds_sub <- t(scale(t(dds_sub)))

  # Heatmap
  p <- pheatmap(
    dds_sub,
    cluster_rows = FALSE,
    cluster_cols = TRUE,
    show_rownames = TRUE,
    fontsize = 10,
    fontsize_row = 12,
    fontsize_col = 12,
    border_color = NA,
    ##cellwidth = 15,
    #cellheight = 10
  )
  p
}

# Create top genes heatmap
p_top <- make_top_genes_heatmap(
  res,
  dds,
  n = TOP_GENES_N,
  output_path = file.path(OUTPUT_DIR, "heatmap_top_genes.pdf")
)
```
## 8. PCA Plot

```{r pca, fig.width = 8, fig.height = 6}
# PCA visualization
plot_pca <- function(dds, filter_genes = TRUE) {
  # Optionally filter out NA symbols and pattern genes
  if (filter_genes) {
    idx <- !is.na(rowData(dds)$symbol)

    if (FILTER_PATTERN != "") {
      idx <- idx & !grepl(FILTER_PATTERN, rowData(dds)$symbol, ignore.case = TRUE)
    }

    dds_filt <- dds[idx, ]
    message("PCA with", nrow(dds_filt), "genes (after filtering)")
  } else {
    dds_filt <- dds
  }

  # VST transform
  vsd <- vst(dds_filt)

  # Extract PCA data
  pca_data <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
  percent_var <- round(100 * attr(pca_data, "percentVar"))

  # Plot with ggplot
  p <- ggplot(pca_data, aes(PC1, PC2, color = condition, label = name)) +
    geom_point(size = 4, alpha = 0.8) +
    ggrepel::geom_text_repel(size = 3.5, max.overlaps = Inf) +
    #geom_text(vjust = -0.7, size = 3.5) +
    expand_limits(
      x = c(min(pca_data$PC1) - 5, max(pca_data$PC1) + 5),
      y = c(min(pca_data$PC2) - 5, max(pca_data$PC2) + 5)
    ) +
    theme_bw(base_size = 14) +
    theme(
      panel.grid = element_blank(),
      legend.position = "right"
    ) +
    xlab(paste0("PC1: ", percent_var[1], "% variance")) +
    ylab(paste0("PC2: ", percent_var[2], "% variance"))
  p
}

# Plot PCA (with filtering)
p_pca_filt <- plot_pca(dds, filter_genes = TRUE)
p_pca_filt

# Also save
ggsave(file.path(OUTPUT_DIR, "pca_plot.pdf"), p_pca_filt, width = 8, height = 6)
```

```
renv::snapshot()
```